---
title: Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory
  Concentrations
author:
- affilnum: 1
  email: hollister.jeff@epa.gov
  name: Jeffrey W. Hollister
- affilnum: 1
  name: Betty J. Kreakie
affiliation:
- affil: US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA
  affilnum: 1
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  html_document: null
  word_document: null
capsize: normalsize
csl: components/ecology.csl
documentclass: article
fontsize: 11pt
linenumbers: no
bibliography: components/manuscript.bib
spacing: doublespacing
abstract: no
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Microsystin Chlorophyll Manusript}
-->

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
if(!require(microcystinchla)){
  devtools::install_github("jhollist/microcystinchla",
                           auth_token = getOption("Github_Access"))
}
if(!require(viridis)){
  install.pacakges("viridis") 
}
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
library("e1071")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

# Table Captions from @DeanK on http://stackoverflow.com/questions/15258233/
# using-table-caption-on-r-markdown-file-using-knitr-to-use-in-pandoc-to-convert-t
# Figure captions are handled by LaTeX

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```

```{r analysis , include=FALSE, echo=FALSE, cache=TRUE}
#All analysis in here, that way all bits of the paper have access to the final objects
#Place tables and figures and numerical results where they need to go.



################################################################################
# Data Prep
################################################################################
data_url<-"http://water.epa.gov/type/lakes/assessmonitor/lakessurvey/upload/NLA2007_Recreational_ConditionEstimates_20091123.csv"
nla_dat <- get_nla(data_url)
nla_dat <- mutate(nla_dat,log_chla = log10(CHLA),log_micro = log10(MCYST_TL_UGL))

################################################################################
# Condtional Probability
################################################################################
mc_guides <- c(0.3, 1, 1.6, 2, 4, 10, 20)
mc_names <- c("U.S. EPA","WHO","U.S. EPA","WHO","WHO","WHO","WHO")
mc_type <- c("Drinking","Drinking","Drinking","Recreational","Recreational","Recreational","Recreational")
mc_chla <- vector("numeric",length(mc_guides))
#wgt <- nla_dat$WGT_NLA
wgt<- rep(1,nrow(nla_dat))

epa_child_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[1]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100, xW=wgt)

mc_chla[1] <- get_chla(epa_child_cp,0.5)

who_drink_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[2]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100, xW=wgt)

mc_chla[2] <- get_chla(who_drink_cp,0.5)

epa_adult_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[3]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100, xW=wgt)

mc_chla[3] <- get_chla(epa_adult_cp,0.5)

who_rec_low1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log10(mc_guides[4]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100, xW=wgt)

mc_chla[4] <- get_chla(who_rec_low1_cp,0.5)

#Agreement Tables
child_conmat<-conmat_to_df(table(chla=nla_dat$CHLA>=mc_chla[1],
                                 mcsyt=nla_dat$MCYST_TL_UGL>=0.3))
who_drink_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[2],
                        nla_dat$MCYST_TL_UGL>=1))
adult_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[3],
                    nla_dat$MCYST_TL_UGL>=1.6))
who_rec_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[4],
                      nla_dat$MCYST_TL_UGL>=2))
total_accuracy<-NULL
total_accuracy$epa_child<-round((child_conmat[1,1]+child_conmat[2,2])/
  sum(child_conmat),2)
total_accuracy$who_drink<-round((who_drink_conmat[1,1]+who_drink_conmat[2,2])/
  sum(who_drink_conmat),2)
total_accuracy$epa_adult<-round((adult_conmat[1,1]+adult_conmat[2,2])/
  sum(adult_conmat),2)
total_accuracy$who_rec<-round((who_rec_conmat[1,1]+who_rec_conmat[2,2])/
  sum(who_rec_conmat),2)

fn_percent<-NULL
fn_percent$epa_child<-round((child_conmat[1,2])/
  sum(child_conmat),2)
fn_percent$who_drink<-round((who_drink_conmat[1,2])/
  sum(who_drink_conmat),2)
fn_percent$epa_adult<-round((adult_conmat[1,2])/
  sum(adult_conmat),2)
fn_percent$who_rec<-round((who_rec_conmat[1,2])/
  sum(who_rec_conmat),2)

```


\singlespace

\vspace{2mm}\hrule

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and require advanced understanding to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper we use this association to provide estimates of chlorophyll *a* that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach that has been used in other water quality settings, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For microcystin concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.   

\vspace{3mm}\hrule
\doublespace

#Introduction

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014], and the US Environmental Protection Agency (USEPA) has recently released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.  

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing remains the gold standard for quantifying microcystin-LR concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are not yet widely used and are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).    Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and understood components of water quality.

Chlorophyll *a* is a very commonly measured component of water quality that is also known to be associated with Microsystin-LR concentrations [@lee2000variation; @paerl2013harmful; heisler2008eutrophication].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that measure chlorohpyll *a*.   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utilize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

#Methods

##Data
We used the 2007 National Lakes Assessment (NLA) water quality and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 and data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.  

##Conditional Probability Analysis
We used a conditional probability analysis (CPA) approach to explore associations between chlorophyll *a* concentrations and World Health Organization (WHO) and U.S. Environmental Protection Agency (U.S. EPA) microcystin-LR health advisory levels [@paul2011probability].  Many levels have been suggested (Table \ref{tab:microcystin_levels}), but lakes with higher microcystin-LR concentrations in the NLA were rare. Only `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat)),2)` % of lakes sampled had a concentration greater than 10.  Thus, for this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were  0.3, 1, 1.6, and 2 $\mu$g/L.

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with an upper confidence level of 50% or greater.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

#Results

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)` and the mean was `r mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)`.  Lastly, of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)`.  The associations between chlorophyll *a* and the upper confidence interval with a conditional probability of 50% were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).   


#Discussion
The association between Log 10 microcystin-LR and Log 10 chlorophyll *a* show a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain concentration.  

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  

Furthermore, the chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin-LR health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin-LR exceedances were `r total_accuracy$epa_child*100`% for the USEPA children's advisory, `r total_accuracy$who_drink*100`% for the WHO drinking water advisory,  `r total_accuracy$epa_adult*100`% for the USEPA adult advisory and, `r total_accuracy$who_rec*100`% for the WHO regreational advisory (Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table}, \ref{tab:adult_conmat_table}, & \ref{tab:who_rec_conmat_table}).  But Total Accuracy is one part of the prediction performace with which we are concerned.  

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  This is an acceptable rate, especially when no other tests are available for a given water body. 

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.  

A second potential use is to identify possible bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.  

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous meausrement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases. 

\newpage

#Figures

```{r chla_micro_scatter, echo=FALSE, fig.width=6, fig.cap="Scatterplot showing association betweeen chlorophyll \\textit{a} and microcystin-LR. \\label{fig:chla_micro_scatter}"}
plot_scatter(nla_dat,xvar="log_chla",yvar = "log_micro",pt_col = "black",
             x=expression(paste('Log10 Chl ', italic("a"),' (',mu,'g/L)')),
             y=expression(paste('Log10 microcystin-LR',' (',mu,'g/L)')))
```

\newpage

<!--
```{r chla_micro_cdf, echo=FALSE, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. \\label{fig:chla_micro_cdf}"}
plot_cdf(nla_dat[,c("log_chla","log_micro")],"")
```

\newpage-->

```{r epa_child_cp_plot, echo=FALSE, fig.width=6, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. A.) Plot for U.S. EPA childrens drinking water advisory (0.3 $\\mu$g/L). B.) Plot for WHO drinking water advisory (1 $\\mu$g/L). C.) Plot for U.S. EPA adult drinking water advisory (1.6 $\\mu$g/L). D.) Plot for WHO recreational advisory (2 $\\mu$g/L). \\label{fig:multi_cp_plot}"}
multiplot(
  plot_cp(epa_child_cp,x = "Log 10 Chlorophyll", 
          y = "Conditional Probability",
          title="A."),
  plot_cp(who_drink_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="B."),
  plot_cp(epa_adult_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="C."),
  plot_cp(who_rec_low1_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="D."),
  cols=2
)
```

\newpage

#Tables

```{r microcystin_levels,echo=FALSE,results='asis', tab.cap="Various suggested microcystin-LR health advisory concentrations. \\label{tab:microcystin_levels}"}
source_of_stnd <- c("WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "U.S. EPA",
                    "U.S. EPA")
type_of_stnd <- c("Recreational",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Drinking",
                  "Drinking",
                  "Drinking")
concentration <- c("2-4 $\\mu$g/L",
                   "10-20 $\\mu$g/L",
                   "20-2000 $\\mu$g/L",
                   ">2000 $\\mu$g/L",
                   "1 $\\mu$g/L",
                   "0.3 $\\mu$g/L",
                   "1.6 $\\mu$g/L")
#note <- c("Low probability of adverse health effects",
#          "Moderate probability of adverse health effects",
#          "High probability of adverse health effects, These higher #concentrations typically only seen in visible scums/algal mats",
#          "Very high probability of adverse health effects",
#          "",
#          "Children younger than school age",
#          "Individuals School age and older")
mc_conc_table <- data.frame(source_of_stnd, type_of_stnd, concentration) %>%
  arrange(type_of_stnd)
names(mc_conc_table)<-c("Source","Type","Concentration")
#pandoc.table(mc_conc_table,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_conc_table)
```

\newpage

```{r mc_chla_table, results='asis', echo=FALSE, tab.cap="Chlorophyll \\textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
#pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_chla_tbl)
```

\newpage

```{r child_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the U.S. EPA childrens drinking water advisory. \\label{tab:child_conmat_table}"}
knitr::kable(child_conmat)
```

\newpage

```{r who_drink_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the WHO drinking water advisory. \\label{tab:who_drink_conmat_table}"}
knitr::kable(who_drink_conmat)
```

\newpage

```{r adult_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the U.S. EPA adult drinking water advisory. \\label{tab:adult_conmat_table}"}
knitr::kable(adult_conmat)
```

\newpage

```{r who_rec_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the WHO recreational water advisory. \\label{tab:who_rec_conmat_table}"}
knitr::kable(who_rec_conmat)
```

\newpage

#References