---
title: 'Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory Concentrations'
author:  
- name: Jeffrey W. Hollister
  affilnum: 1
  email: hollister.jeff@epa.gov
- name:  Betty J. Kreakie
  affilnum: 1
- name: Dorothy Q. Kellog
  affilnum: 2
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
- affilnum: 2
  affil: University of Rhode Island, Department of Natural Resrouces Science, Kingston, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: no
bibliography: components/manuscript.bib
abstract: no
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Microsystin Chlorophyll Manusript}
-->

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
devtools::install_github("jhollist/microcystinchla",
                         auth_token = getOption("Github_Access"))
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

# Table Captions from @DeanK on http://stackoverflow.com/questions/15258233/
# using-table-caption-on-r-markdown-file-using-knitr-to-use-in-pandoc-to-convert-t
# Figure captions are handled by LaTeX

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```

```{r analysis , include=FALSE, echo=FALSE, cache=TRUE}
#All analysis in here, that way all bits of the paper have access to the final objects
#Place tables and figures and numerical results where they need to go.



################################################################################
# Data Prep
################################################################################
data_url<-"http://water.epa.gov/type/lakes/assessmonitor/lakessurvey/upload/NLA2007_Recreational_ConditionEstimates_20091123.csv"
nla_dat <- get_nla(data_url)
nla_dat <- mutate(nla_dat,log_chla = log(CHLA),log_micro = log(MCYST_TL_UGL))

################################################################################
# Condtional Probability
################################################################################
mc_guides <- c(0.04, 0.3, 0.8, 1, 1.6, 4, 6, 10, 14, 20)
mc_names <- c("MN","EPA_Child","CA","WHO","EPA_Adult","WHO","Multiple","Multiple","Mulitple","Multiple")
mn_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[1]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
epa_child_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[2]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
ca_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[3]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
who_drink_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[4]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
epa_adult_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[5]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
who_rec_low_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[6]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
mult_state_rec_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[7]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
il_org_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[8]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
ma_ri_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[9]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)
who_rec_med_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
                     xImpair = log(mc_guides[10]), ProbComp = "gt", Exceed = "gte",
                     ci = TRUE, R = 100)

```


\singlespace

\vspace{2mm}\hrule

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->


\vspace{3mm}\hrule
\doublespace

#Introduction

#Methods

```{r microcystin_levels,echo=FALSE,results='asis'}
source_of_stnd <- c("WHO, IN",
                    "WHO, IN, KS",
                    "WHO, IN, IA, KS, NE, OH",
                    "WHO",
                    "WHO, OH, OR",
                    "MN",
                    "CA",
                    "IL, OR",
                    "MA, RI",
                    "OH, VT, VA, WA",
                    "U.S. EPA",
                    "U.S. EPA")
type_of_stnd <- c("Recreational",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Drinking",
                  "Drinking",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Drinking",
                  "Drinking")
concentration <- c("2-4 ug/l",
                   "10-20 ug/l",
                   "20-2000 ug/l",
                   ">2000 ug/l",
                   "1 ug/l",
                   "0.04 ug/l",
                   "0.8 ug/l",
                   "10 ug/l",
                   "14 ug/l",
                   "6 ug/l",
                   "0.3 ug/l",
                   "1.6 ug/l")
#note <- c("Low probability of adverse health effects, state advisories",
#          "Moderate probability of adverse health effects, state advisories",
#          "High probability of adverse health effects, state advisories, These higher #concentrations typically only seen in visible scums/algal mats",
#          "Very high probability of adverse health effects, These higher concentrations #typically only seen in visible scums/algal mats",
#          "","","","","","",
#          "Children younger than school age",
#          "Individuals School age and older")
mc_conc_table <- data.frame(source_of_stnd, type_of_stnd, concentration) %>%
  arrange(type_of_stnd)
names(mc_conc_table)<-c("Source","Type","Concentration")
pandoc.table(mc_conc_table,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

Based on the range of microcystin concentrations used in various policy settings, we evaulated associated chlorohpyll *a* concentrations for an effect at 0.04, 0.3, 0.8, 1, 1.6, 4, 6, 10, 14, and 20 ug/l.   

##Data and Study Area

#Results


#Discussion

#Figures


```{r mn_cp_plot,echo=FALSE}
cp_plot(mn_cp)
```

\newpage

```{r epa_child_cp_plot,echo=FALSE}
cp_plot(epa_child_cp)
```


```{r ca_cp_plot,echo=FALSE}
cp_plot(ca_cp)
```

\newpage

```{r who_drink_cp_plot,echo=FALSE}
cp_plot(who_drink_cp)
```

\newpage

```{r epa_adult_cp ,echo=FALSE}
cp_plot(epa_adult_cp)
```

\newpage

```{r who_rec_low_cp, echo=FALSE}
cp_plot(who_rec_low_cp)
```

\newpage

```{r mult_state_rec_cp,  echo=FALSE}
cp_plot(mult_state_rec_cp)
```

\newpage

```{r il_org_cp, echo=FALSE}
cp_plot(il_org_cp)
```

\newpage

```{r ma_ri_cp, echo=FALSE}
cp_plot(ma_ri_cp)
```

\newpage

```{r who_rec_med_cp, echo=FALSE}
cp_plot(who_rec_med_cp)
```

\newpage

#References